---
title: "R Notebook"
output:
  html_document:
    df_print: paged
---
Load the libraries
```{r}
library(dplyr)
library(readr)
library(readxl)
```

Read protein expression data from the input file
```{r}
protein_data <- read_xlsx("./data/input/protein_data_avgN.xlsx")
protein_data <- as.data.frame(protein_data, check.names = FALSE)
rownames(protein_data) <- protein_data$protein
#remove $protein column
protein_data <- protein_data[,-1]
#round up the data keeping only 2 decimal places
protein_data <- round(protein_data, 2)
```

```{r}
rna_data <- read_csv("./browse/mirna_data_avgN.csv")
rna_data <- as.data.frame(rna_data, check.names = FALSE)
rownames(rna_data) <- rna_data$rna
#remove $protein column
rna_data <- rna_data[,-1]
rna_data <- round(rna_data, 2)
#round up the data keeping only 2 decimal places

data<-lipid_data
```

Lipid data
```{r}
lipid_data <- read.csv("./browse/lipid_data_avgN.csv")
lipid_data <- as.data.frame(lipid_data, check.names = FALSE)
rownames(lipid_data) <- lipid_data$lipid
#remove $protein column
lipid_data <- lipid_data[,-1]
lipid_data <- round(lipid_data, 2)
data <- lipid_data
```
```{r}
  group1 <- c("3DEV")
  group2 <- c("2DSuper", "3DSuper", "FPLCSuper")
  group3 <- c("2DExomere", "3DExomere", "FPLCExomere")
  group4 <- c("2DEVP", "3DEVP")
  
  groups <- list(EVs = group1, Super = group2, Exomere = group3, EVp = group4)
  
  # Extract the data for the selected lipid
  lipid_data_subset <- data[input$lipid, , drop = FALSE]
  
  # Initialize vectors for group names and expression values
  group_names <- c()
  expression_values <- c()
  
  # Iterate over the groups and extract data
  for (group_name in names(groups)) {
    group_cols <- groups[[group_name]]
    valid_cols <- intersect(group_cols, colnames(lipid_data_subset))
    
    if (length(valid_cols) > 0) {
      group_names <- c(group_names, rep(group_name, length(valid_cols)))
      expression_values <- c(expression_values, as.vector(t(lipid_data_subset[, valid_cols])))
    } else {
      cat("Warning: No valid columns found for group", group_name, "\n")
    }
  }
  
  # Create a data frame for plotting
  grouped_data <- data.frame(
    Group = factor(group_names, levels = c("EVs", "Super", "Exomere", "EVp")),
    Expression = expression_values
  )
  
  # Debugging: Print the structure of grouped_data
  print(head(grouped_data))
  
  # Generate the plot using pirateplot
  pirateplot(formula = Expression ~ Group, data = grouped_data,
             theme = "white", 
             pal = c("EVs" = "steelblue", "Super" = "lightcoral", "Exomere" = "mediumseagreen", "EVp" = "goldenrod"),
             ylab = "Expression", 
             xlab = "Group", 
             main = paste("Grouped lipid presence for", input$lipid))
```

Initialize an empty data frame for storing GO protein data
```{r}
GOprotein_data <- data.frame(matrix(ncol = ncol(protein_data), nrow = 0), check.names = FALSE)
condition_list <- as.character(colnames(protein_data))
colnames(GOprotein_data) <- condition_list

```

Get the list of .tsv files from the directory, iterate through each file, read the tsv file, split the protein names in 'matching proteins in your input (labels)' column, ensure the condition column exists, iterate through each protein, ensure the protein row exists, ensure that the expression value is correctly assigned, and check the resulting GOprotein_data

```{r}
# Get the list of .tsv files from the directory
file_list <- list.files(path = "./data/input/ranked", pattern = "\\.tsv$", full.names = TRUE)

# Iterate through each file
for (file in file_list) {
  
  # Extract the condition name (XX) from the file name
  condition <- gsub("^(.*?)_enrichment\\..*", "\\1", basename(file))
  print(paste("Processing condition:", condition))
  
  # Read the tsv file
  tsv_data <- read_tsv(file, show_col_types = FALSE)
  
  # Split the protein names in 'matching proteins in your input (labels)' column
  protein_list <- unlist(strsplit(tsv_data$`matching proteins in your input (labels)`, ","))
  protein_list <- as.character(protein_list)
  #print(paste("Proteins found:", paste(protein_list, collapse = ", ")))
  
  # Ensure the condition column exists
  if (!(condition %in% colnames(GOprotein_data))) {
    GOprotein_data[[condition]] <- NA
  }
  
  # Iterate through each protein
  for (protein in protein_list) {
    #print(paste("Processing protein:", protein))
    
    # Ensure the protein row exists
    if (!(protein %in% rownames(GOprotein_data))) {
      GOprotein_data <- rbind(GOprotein_data, setNames(as.list(rep(NA, ncol(GOprotein_data))), colnames(GOprotein_data)))
      rownames(GOprotein_data)[nrow(GOprotein_data)] <- protein
    }
    
    # Ensure that the expression value is correctly assigned
    if(protein %in% rownames(protein_data)) {
      GOprotein_data[protein, condition] <- protein_data[protein, condition]
    } else {
      warning(paste("Protein", protein, "not found in protein_data"))
    }
  }
}

# Check the resulting GOprotein_data
#print(GOprotein_data)

```
Remove columns with all NA values

```{r}
#remove columns with all NA values
GOprotein_data <- GOprotein_data[, colSums(is.na(GOprotein_data)) < nrow(GOprotein_data)]
head(GOprotein_data)
```
Based on the GOprotein_data, create a heatmap and PCA plot
```{r}
heatmap_data <- as.data.frame(GOprotein_data, check.names = FALSE)
heatmap_data[is.na(heatmap_data)] <- 0
heatmap_data <- heatmap_data[order(rowMeans(heatmap_data, na.rm = TRUE), decreasing = TRUE),]
heatmap_data <- heatmap_data[,order(colMeans(heatmap_data, na.rm = TRUE), decreasing = TRUE)]
```

Heatmap
```{r}
library(pheatmap)
heatmap_data2 <- heatmap_data
heatmap_data2 <- heatmap_data2[, !colnames(heatmap_data2) %in% '3DhEV']
pheatmap(heatmap_data2, scale = "column", cluster_rows = TRUE, cluster_cols = TRUE, show_rownames = FALSE, show_colnames = TRUE, color = colorRampPalette(c("blue", "white", "red"))(100))
```

PCA plot libraries
```{r}
library(ggplot2)
library(ggrepel)
library(ggfortify)
```

PCA plot
```{r}
# Ensure heatmap_data is numeric
PCA_data <- as.data.frame(lapply(heatmap_data, as.numeric, check.names = FALSE))
PCA_data <- PCA_data[, apply(PCA_data, 2, var, na.rm = TRUE) != 0]


# Perform PCA without additional scaling
pca_data <- prcomp(t(PCA_data), scale. = FALSE)

# Assuming the condition corresponds to the columns in the original data
condition <- colnames(GOprotein_data)

# Create a data frame for plotting
pca_df <- data.frame(pca_data$x, condition = condition)

library(ggplot2)
library(ggrepel)

# Plot the PCA with repelled labels
ggplot(pca_df, aes(x = PC1, y = PC2, label = condition, color = condition)) +
    geom_point(size = 3) +
    geom_text_repel(size = 3, max.overlaps = 10, box.padding = 0.5, point.padding = 0.5, segment.color = 'grey50') +
    theme_minimal() +
    theme(legend.position = "right") +
    labs(x = paste0("PC1 (", round(pca_data$sdev[1]^2 / sum(pca_data$sdev^2) * 100, 2), "%)"),
         y = paste0("PC2 (", round(pca_data$sdev[2]^2 / sum(pca_data$sdev^2) * 100, 2), "%)") #,
         #title = "PCA Plot")


```
